Scalar Feedback based Joint Time-Frequency Precoder interpolation for MIMO-OFDM Systems

### 2019-12-02 paper/main.tex
Minor changes

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency. While these
methods have been shown to be very effective for precoder feedback,
they involve operations over higher dimensional manifolds. In
particular, optimization and interpolation is complicated, especially
for manifolds where geodesic paths for interpolation may be difficult
to obtain, especially for joint time-frequency predictive coding. An
alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices parameterized into
independent scalar parameters that correspond to Givens rotations and
Household Transformations of the unitary precoding matrices
\cite{4114278,4556174}. These approaches have been combined with
adaptive delta modulation effectively track precoding matrices both in
the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately.

@@

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$ could be modelled as a flat-fading MIMO channel. This flat-fading MIMO channel can be modelled as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier.

@@

$N_T$ rows and $N_R$ columns can be decomposed as follows~\cite{4114278} (we suppress
the $(i, t)$ subscripts in the decomposition matrices)

@@

that $\phi_k \in (-\pi, \pi]$ is uniformly distributed between the
two extremes~\cite{4114278}, while $\theta_l$ is distributed as~\cite{4114278}

@@

\caption{\label{fig:adpm-fig}Subcarriers which need to be quantized}

@@

One issue that is associated with the tracking the $\phi_i$ parameters
is that they may change abruptly between $-\pi$ and
$\pi$ due to jumps of $2\pi$. We have avoided this by unwrapping the
phases to facilitate continuous tracking.
\subsection{Joint Time Frequency Interpolation}
\label{interp}
After obtaining the $\beta$ values (using the single bit feedback) at
the transmitter, we can construct the precoder at subcarrier indices
$pk$ or $pk+q$. To construct the rest of the subcarriers, we will use
precoder parameters from both the past and future and time instances
to do the interpolation (using backtracking to ensure causality). In
Fig.~\ref{fig:adpm-fig}, each parameter is interpolated using the
neighbouring subcarriers using bilinear interpolation (equivalent to
linear interpolation in two dimensions). These interpolated values and
then used to reconstruct the precoding matrices all together. Thus,
the approach here is able to exploit the time and frequency
correlation jointly to enhance performance.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model for
both the ITU Vehicular-A channel and ITU Pedestrian-A channel power
delay profiles~\cite{recommendation1997guidelines}. The BER achieved by the proposed predictive
quantization based precoder is compared against the ideal precoder for
100 channel evolutions in each run and averaged. The normalised
Doppler is assumed to be $f_dT_s = 3.5\times10^{-4}$, with symbol time
period $T_s = 5\times10^{-8} s$. For both the Vehicular and Pedestrian
channel profiles, we consider $N_T=4$ and $N_R=2$. Therefore, the
total number of parameters that determine the precoder are
$N_{R}(2N_{T} - N_{R}) = 12$, with 5$\theta_l$s and
7$\phi_i$s. For the Vehicular channel profile, we assume that the
number of subcarriers is 1024, while for the Pedestrian channel
profile it is 64. We take the spacing between fed back subcarrier
indices $p$ to be $33$ and $9$ for the Vehicular
and Pedestrian channel profiles respectively (i.e. for the Vehicular
profile, the subcarriers indices fed back would be $0, 33, 66, \ldots
1023$ and for the Pedestrian, $0, 9, 18, \ldots 63$).

>>>>

The problem of effective quantization and feedback of precoders for
MIMO-OFDM systems has attracted significant attention in recent
years. Many approaches involve codebook design over the Grassmann
manifold~\cite{mondal2007quantization,schwarz2013adaptive,5946308},
the Stiefel manifold \cite{6891198,Gupt1905:Predictive} and the Flag
manifold~\cite{pitaval2013codebooks}, wherein the manifold structure
of the precoders is used to obtain tangents for predictive
quantization and interpolation across time and frequency. While these
methods have been shown to be very effective for precoder feedback,
they involve operations over higher dimensional manifolds. In
particular, optimization and interpolation is complicated, especially
for manifolds where geodesic paths for interpolation may be difficult
to obtain, especially for joint time-frequency predictive coding. An
alternate approach is to parametrize the precoding matrices using
scalar parameters. The space of unitary matrices parameterized into
independent scalar parameters that correspond to Givens rotations and
Householder transformations of the unitary precoding matrices
\cite{4114278,4556174}. These approaches have been combined with
adaptive delta modulation effectively track precoding matrices both in
the time~\cite{4114278} and frequency domains~\cite{4556174}, albeit
separately.

@@

\section{System Model}
For a MIMO-OFDM system with $N_T$ transmitter and $N_R$ receiver
antennas, the channel for the $i$-th subcarrier at time instant $t$
can be modelled as a flat-fading MIMO channel. This flat-fading MIMO
channel can be modelled as
\label{section2}
\begin{equation}
\by_{i,t} = \bH_{i,t}\tilde{\bV}_{i,t} \bx_{i,t}+ \bf{\eta_{i,t}}
\end{equation}
where $\bH_{i,t} \in \mathbb{C}^{N_R \times N_T}$ is the channel
matrix, $\bx_{i,t} \in \mathbb{C}^{N_d}$ is the input data vector,
$\by_{i,t} \in \mathbb{C}^{N_R}$ is the output vector,
$\tilde{\bV}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ is the unitary
precoder, and $\eta$ is the complex additive white Gaussian noise
distributed as ${\mathcal{CN}}(0,I_{N_R})$. We consider the case where
$N_R < N_T$, and the data vector size $N_d = N_R$.  The ``thin'' SVD
(that keeps only those right and left singular vectors that correspond
to non-zero singular values) of $\bH$ is given by
$\bH_{i,t} = \bU_{i,t} \Sigma_{i,t} \bV_{i,t}^{H}$, where
$\bU_{i,t} \in \mathbb{C}^{N_R \times N_R}$,
$\Sigma_{i,t} \in \mathbb{C}^{N_R \times N_R}$ and
$\bV_{i,t} \in \mathbb{C}^{N_T \times N_R}$. $\bU_{i,t}$ and
$\bV_{i,t}$ are matrices whose columns are orthonormal, and
$\Sigma_{i,t}$ contains the singular values
$\sigma_1 \geq \ldots \geq \sigma_{N_R} > 0$ of $\bH_{i,t}$. The
entries of $\bH_{i,t}$ are assumed to be distributed
i.i.d. $\\mathcal{CN}(0,1)$. To obtain a precoder at the
transmitter, we need to quantize and feed back information about
$\bV_{i,t}$ for each subcarrier. However, the frequency coherence and
temporal correlations would reduce the overall feedback requirement,
as discussed in subsequent sections.

@@

$N_T$ rows and $N_R$ columns can be decomposed as
follows~\cite{4114278} (we suppress the $(i, t)$ subscripts in the
decomposition matrices)

@@

that $\phi_k \in (-\pi, \pi]$ is uniformly distributed between the two
extremes~\cite{4114278}, while $\theta_l$ is distributed
as~\cite{4114278}

@@

\caption{\label{fig:adpm-fig}Subcarriers which need to be
  quantized. In the figure, we see that the subcarrier indicated by
  red arrows has a precoder whose parameters are determined jointly
  using the neighbouring parameters, both in time and frequency.}

@@

One issue that is associated with the tracking the $\phi_i$ parameters
is that they may change abruptly between $-\pi$ and
$\pi$ due to jumps of $2\pi$. We have avoided this by unwrapping the
phases to facilitate continuous tracking (i.e., we avoid jumps in
phase of magnitude close to $2\pi$).
\subsection{Joint Time Frequency Interpolation}
\label{interp}
After obtaining the $\beta$ values (using the single bit feedback) at
the transmitter, we can construct the precoder at subcarrier indices
$pk$ or $pk+q$, in keeping with the same notation as the previous
section. To construct the rest of the subcarriers, we will use
precoder parameters from both the past and future and time instances
to do the interpolation (using backtracking to ensure causality). In
Fig.~\ref{fig:adpm-fig}, each parameter is interpolated using the
neighbouring subcarriers using bilinear interpolation (equivalent to
linear interpolation in two dimensions). These interpolated values and
then used to reconstruct the precoding matrices all together. Thus,
the approach here is able to exploit the time and frequency
correlation jointly to enhance performance.

@@

\section{Simulation and discussion}
\label{section3}
We now present simulations of the scalar feedback based joint
time-frequency predictive quantization outlined in the previous
section. In particular, we analyze the performance of the proposed
quantization and interpolation method for time-varying MIMO-OFDM
systems. The channel is simulated using the typical Jakes Model for
both the ITU Vehicular-A channel and the COST channel power delay
profiles~\cite{recommendation1997guidelines,molisch2006cost259,cost1989cost}. The
BER achieved by the proposed predictive quantization based precoder is
compared against the ideal precoder for 100 channel evolutions in each
run and averaged. The normalised Doppler is assumed to be
$f_dT_s = 3.5\times10^{-4}$, with symbol time period
$T_s = 5\times10^{-8} s$. For both the Vehicular and Pedestrian
channel profiles, we consider $N_T=4$ and $N_R=2$. Therefore, the
total number of parameters that determine the precoder is
$N_{R}(2N_{T} - N_{R}) = 12$, with five $\theta_l$s and and seven
$\phi_i$s. For the fast fading channels, we assume that the number of
subcarriers is 1024, while for the slower varying channel profiles, we
take 64 subcarriers. We assume that the spacing between fed back
subcarrier indices $p$ to be $33$ and $9$ for the fast and slow
channel profiles respectively (i.e. for the slower channels, the
subcarriers indices fed back would be $0, 33, 66, \ldots 1023$ and for
the Pedestrian, $0, 9, 18, \ldots 63$).
